Interactive Visual Analysis of Image-Centric Cohort Study Data

### 2014-06-16 draft.tex
First Version of rewrite of application section

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
% TODO: Self added Packages - need to be removed afterwards
%\usepackage{babel}
\usepackage[normalem]{ulem}
\usepackage{color}
\newcommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
\newcommand{\add}[1]{\textcolor{blue}{\uline{#1}}}

@@

%% Abstract section.
\abstract{%
%(The problem).
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
%
%The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
%
%Modern cohort studies incorporate medical image data.
%
%
%The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
%\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It allows a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are employed to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of people who are randomly selected to avoid any bias.
%
\rem{Due to this random selection, the majority of the cohort does not suffer from a specific health problem.}
\add{The people are selected without focus on a certain disease.}
%
A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%
Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.

@@

\subsection{Epidemiological Data} \label{EpidemiologicalData}
Epidemiological data are highly heterogenous and incomplete.
%
Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yield a complex data space for each subject.
%
For ethical, legal or medical reasons some features cannot be gathered for each subject.
%
The most obvious example are women-specific questions about menstrual status or number of born children.
%
Follow-up examinations or questions about conditions like medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
%

@@

\paragraph{Image acquisition.} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
Therefore, MRI is the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%
%The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
The equipment used to gather medical image data is \rem{kept, if possible, on the initial software and hardware version} \add{not updated or changed} during a longitudinal study to ensure comparability in and between acquisition cycles.

\paragraph{Image analysis.} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
%
Information derived by landmarks, such as top and bottom point of a vertebra, are by far not as expressive and versatile as segmentation masks describing its whole shape.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).

@@

\subsubsection{System Structure} \label{Structure and Workflow}
We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
\begin{itemize}
	\item The \emph{sidebar}, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely.
	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view. This view on the data comes natural to epidemiologists.
	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix.
\end{itemize}
%
% The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
% 
% Features can be dragged from the sidebar into the canvas area.
% %
% Doing so will automatically create an information visualization suitable for the current data type.
% %
% 3D shape information about the investigated structure is displayed for each variable manifestation.
% %
% By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
% %
% Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
% %
% Shape-based clustering can either be applied to all subjects or subgroups.
% 
% All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
% %
% Details to the different views are presented in the following sub-sections.

@@

The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \url{http://getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%
Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
%
The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
%
\texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} allows GPU-accelerated data rendering using \texttt{WebGL}.
%
Communication between client and server is enabled via the WebSockets protocol.
%
Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we had to access them using the \texttt{NodeJS} server.
%
We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it in a proper format to the client.
%
All parameter-steered console applications can be incorporated in this context.
\\\\
\section{Application}
This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
%
We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
%
This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
%
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
% Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
% \begin{itemize}
% 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
% 	\item prognostic features for diagnosis and treatment of lumbar back pain
% 	\item determination of particularly effected risk groups
% \end{itemize}
%
Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.

@@

\section{Summary and Conclusion}
In this paper, we presented an \emph{IVA} framework for the analysis of complex image-centric epidemiological data.
%
Hence, the framework allows for both, hypothesis-driven analysis and hypothesis generation.
%
The visualization of multivariate data using connected views and different views allows to get fast visual feedback about subject groups.
%
Brushing and linking makes the data tangible and adaptable to formulated hypotheses.
%
The use of pivot tables is familiar to epidemiologists while embracing the power of interactive adjustment of the shown features.
%
The automatic suggestion of correlations using contingency methods like \emph{Cram\'{e}r's V} triggers \emph{hypothesis generation} by highlighting features potentially overlooked by the experts.
%
Shape-based clustering assesses the variability of a anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.

>>>>

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
% TODO: Self added Packages - need to be removed afterwards
%\usepackage{babel}
\usepackage[normalem]{ulem}
%\usepackage{color}
\usepackage{xcolor}
\newcommand{\rem}[1]{\textcolor{red}{\sout{#1}}}
\newcommand{\add}[1]{\textcolor{blue}{\uline{#1}}}
\newcommand{\com}[1]{\textcolor{orange}{\uline{#1}}}

@@

%% Abstract section.
\abstract{%
%(The problem).
\com{ToDo: Reviewer-Feedback: The abstract should be formulated more concisely and focus on What? 
Why? How? What follows?}
Epidemiological population studies impose information about a set of subjects (a \emph{cohort}) to characterize disease-specific risk factors.
%
Cohort studies comprise of heterogenous variables (\emph{features}) describing the medical condition as well as demographic and lifestyle factors and more recently, medical image data.
%
%The data are analyzed using a priori defined hypotheses to find statistically significant correlations between features (\emph{associations}).
%
%Modern cohort studies incorporate medical image data.
%
%
%The statistically driven epidemiological workflow only allows to determine \emph{associations} between image-derived metrics, such as distances extracted from landmarks of the segmentation model.
%\\\\
%(New Solution).
We propose an Interactive Visual Analysis (\emph{IVA}) approach that enables epidemiologists to examine both image-based as well as non-image data, e.g., sociodemographic features and attributes derived from the image data.
%
This is achieved by combining brushing and linking enabled coordinated information visualization views and interactive 3D shape renderings with epidemiological data representations such as pivot tables and key figures as association measures.
%(Validation).
The presented concepts are applied expert-guided to gather and evaluate hypotheses about the aging process of the lumbar spine.
%
%(Results). 
It allows a more flexible comparison between image and non-image data.
%
%(Implications).
The new framework enables hypotheses validation and generation by incorporating human pattern recognition as well as data mining methods.
%
Using all reliable information from the image segmentation linked to non-image features aims to unveil \emph{associations} by applying an iterative analysis approach.
} % end of abstract

@@

%% \section{Introduction} %for journal use above \firstsection{..} instead
Epidemiology aims at characterizing health and disease by determining risk factors.
%
Clinical problems, such as the selection of diagnostic tools and efficient treatment, are tackled using results of epidemiological research.
%
Also, the introduction of preventive measures in medicine and beyond is based on epidemiological research, where, for example, subgroups with increased risk are identified \cite{Fletcher2012}.
%
Observations made by clinicians in the daily routine are translated into hypotheses for epidemiological research.
%
These are used to determine environmental and lifestyle factors as well as medical examination results that may influence a disease.
%
Potentially useful data variables (\emph{features}) are gathered using structured interviews and clinical examinations.
%
Methods like regression analysis are employed to check the attribute list for statistical soundness.
%
\\\\
Longitudinal population-based studies, such as the Study of Health in Pomerania (SHIP) \cite{Volzke2011}, gather as much information as possible about a defined sample of people (a \emph{cohort}).
%
The cohort consists of several thousands of people who are randomly selected to avoid any bias.
%
\rem{Due to this random selection, the majority of the cohort does not suffer from a specific health problem.}
\add{The subjects are selected without focus on a certain disease.}
%
A large size of the cohort is therefore essential to investigate differences between healthy and diseased people.
%
% The sample is drawn randomized to avoid selection \cite{Fletcher2012}.
% %
% When for example one investigates risk factors for prostate cancer in male subjects, the outcome is strongly dependent on the age.
% %
% Therefore results need to be age-adjusted to be comparable.
% %
% Confounding variables, are often not obvious at all and characterizing them is already an epidemiological result.
%
Cohort studies often include medical image data.
%
Segmentation of these data would allow for anatomical structure analysis and enable correlation with other features.
%
Semi-automatic techniques are more promising but also challenging, since the employed modalities, such as magnetic resonance imaging (MRI) and ultrasound, are subject to inhomogeneity and noise.
%
Analyzing spatial data with respect to other epidemiological factors requires techniques that reach beyond standard statistical methods.

@@

\section{Epidemiological Data} \label{EpidemiologicalData}
\com{Primary Reviewer Feedback: Another weakness is on the data analysis; while there is information about the data within the text (including 2.2, 4.1, and 5.2) I would like to see a section that focuses on the data and the challenges thereof.
This chapter needs to be enhanced! Maybe own section!}
Epidemiological data are highly heterogenous and incomplete.
%
Information about medical history and examinations, genetic conditions, geographical data, questionnaire results and image data yield a complex data space for each subject.
%
For ethical, legal or medical reasons some features cannot be gathered for each subject.
%
The most obvious example are women-specific questions about menstrual status or number of born children.
%
Follow-up examinations or questions about conditions like medications taken after a diagnosed disease also yield features only available for a small amount of subjects.
%

@@

\subsection{Image acquisition} \label{ImageAcquisition} Imaging techniques involving ionizing radiation for the subject are not suitable for ethical reasons.
%
Therefore, MRI is the main method for collecting cohort study imaging data.
%
The image quality is a tradeoff between accuracy and affordability \cite{Preim2014}.
%
This often yields image resolutions inferior to those of clinical day-to-day practice, which makes their analysis more challenging.
%
%The equipment used to gather medical image data is not updated or changed during a longitudinal study to ensure comparability in and between acquisition cycles.
The equipment used to gather medical image data is \rem{kept, if possible, on the initial software and hardware version} \add{not updated or changed} during a longitudinal study to ensure comparability in and between acquisition cycles.

\subsection{Image analysis} \label{ImageAnalysis} Decisions have to be made on how image data are \emph{compared} and \emph{quantified}.
%
Segmentation masks representing the voxels of an anatomical structure would be ideal, since many different key figures, e.g., volume, largest diameter or aspect ratio, can be derived from them.
%
Since reliable and efficient segmentation techniques for these data are not available in general, epidemiologists are forced to measure the data by hand, which is a very tedious work with respect to the number of necessary landmarks and the number of subjects.
%
Information derived by landmarks, such as top and bottom point of a vertebra, are by far not as expressive and versatile as segmentation masks describing its whole shape.
%
They are also prone to a high inter-observer variability and are hard to reproduce.
%
This gains even more importance when analyzing multiple time steps.
%
Morphometric information from landmarks comprises thickness, diameter or length of a structure as well as grey value distribution in an area (used for determining the type of tissue).

@@

\section{System Structure and Design} \label{Structure and Workflow}
\com{Primary Reviewer Feedback: There is little detail about the design, and how it was decided to create this particular design; what interaction was made with the users, how did they influence this design?}
% \subsubsection{System Structure}
We divide the workspace into four major parts, as illustrated in Figure~\ref{fig:visualization} and \ref{fig:similarity}.
\begin{itemize}
	\item The \emph{sidebar}, which contains all epidemiological features. Cluster results are treated like features and are part of the sidebar as well.
	\item The \emph{canvas} holding all visualizations. Elements can be added, arranged, resized and removed freely.
	\item The interactive \emph{pivot table} gives detailed numerical information of the features in the canvas view. This view on the data comes natural to epidemiologists.
	\item The \emph{contingency view} depicts relations for features in the canvas in an adjacency matrix.
\end{itemize}
%
% The sidebar holds all available variables as well as groups either derived by user input or automatic clustering.
% 
% Features can be dragged from the sidebar into the canvas area.
% %
% Doing so will automatically create an information visualization suitable for the current data type.
% %
% 3D shape information about the investigated structure is displayed for each variable manifestation.
% %
% By dropping variables on existing visualizations the system creates a visualization that allows for comparison (e.g. mosaic plots for ordinal variables).
% %
% Elements can be brushed in the information visualizations and are linked to the other representations in the canvas.
% %
% Shape-based clustering can either be applied to all subjects or subgroups.
% 
% All elements in the canvas view are also represented in the interactive Pivot Table which gives detailed information how the subjects are distributed given the displayed variables.
% %
% Details to the different views are presented in the following sub-sections.

@@

The front-end is created using \texttt{Bootstrap}\footnote{Developed by Twitter, \url{http://getbootstrap.com}} as foundation for the layout and basic UI elements using \texttt{HTML5}, \texttt{CSS3} and \texttt{Javascript}.
%
Information visualizations such as scatterplots and bar charts are created using the popular \texttt{Data-Driven Documents (D3.js)} library \cite{D3}, which works well for attaching data to visible elements like vector graphics and provides powerful transformation and mapping tools.
%
The pivot table implementation was adapted using \texttt{PivotTable.js}\footnote{Developed by Nicolas Kruchten, \url{http://nicolas.kruchten.com/pivottable}}.
%
\texttt{Three.js}\footnote{Originally developed by Ricardo Cabello, \url{http://threejs.org}} allows GPU-accelerated data rendering using \texttt{WebGL}.
%
Communication between client and server is enabled via the WebSockets protocol.
%
Since our clustering algorithms are written in \texttt{MatLab}\footnote{Owned by The MathWorks, \url{http://mathworks.com}}, we had to access them using the \texttt{NodeJS} server.
%
We accomplished this by converting it to a parameterized standalone console application that is spawned by \texttt{NodeJS} on client request and then reads the result from the console standard out and returns it in a proper format to the client.
%
All parameter-steered console applications can be incorporated in this context.
\\\\
\section{Application}
\add{This Chapter needs to be reworked completely, since it was subject of the reviewer criticism.}
This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
%
We conducted a case study with two domain experts on a data set compiled to analyze lower back pain. 
%
This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
%
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
%
Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.
%
\subsection{The Lumbar Spine Data Set}
We divide the data set into image and non-image data.
%
There are 127 features describing diagnosed diseases, lifestyle factors, women-specific factors, pain indicators, laboratory values and somatometric features.
%
The image data was acquired on a 1.5~Tesla scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany) by four trained technicians in a standardized way.
%
%The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence (676 / 12 [repetition time msec / echo time msec]; 150$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence (3760 / 106 [repetition time msec / echo time msec]; 180$^\circ$ flip angle; 500~mm field of view; $1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
The spine protocol consisted of sagittal T1- and T2-weighted turbo-spin-echo sequences with rather large slice distance only sufficient for characterizing the lumbar spine \cite{Hegenscheid2013}.

\subsubsection{Data Preprocessing} \label{application:Data Preprocessing}
The data are preprocessed as described in Section~\ref{Data Preprocessing}.
%
\paragraph{Non-Image Data.}

\subsubsection{Shape Visualization and Clustering}
%
The tetrahedron-based detection model described in Section~\ref{application:Data Preprocessing} consists of corresponding grid points for each structure instance.
%
This allows for calculation of shape distance and similarity.
%
This information is used to calculate mean shapes as described in Section~\ref{Interaction- and Visualization Techniques}.

Shape distance is mapped onto color.
%
For dichotomous variables, the color codes distances between mean shapes of the two groups, for variables with more than two manifestations it encodes the distance to the global mean shape of all subjects (see Figure~\ref{fig:hypothesisfree}).
\\\\
Shape-based clustering is carried out via agglomerative hierarchical clustering of the spine canal centerlines (recall Section~\ref{application:Data Preprocessing} and \cite{Klemm2013VMV}).
%
Since the ``correct'' number of clusters in a given group is unknown, an estimate is computed by means of the knee/elbow method \cite{Salvador2004}.
%
The method has proven to produce sound results on a preliminary data set and was able to reproduce textbook knowledge \cite{Klemm2013VMV}.
\subsection{Domain Experts, Setup and Procedure}
%
%Lam et al. divide evaluations of Visual Analysis systems in two categories, 
%
\com{Anmerkung von Steffen einbauen: Evaluierung geht um fachliches Feedback - Wissen über epidemiologie UND rücken und wissen um die Daten - deswegen schwer ganz viele Experten ranzuziehen}
Guided by the work of Lam et al \cite{Lam2012}, we conducted an investigation of \emph{Visual Data Analysis and Reasoning} (VDAR).
%
This approach aims to characterizes the ability of a system to explore data, discover knowledge, generate hypotheses and help formulating decisions.
%
Since it is hard to quantify when these goals are achieved, Lam et al. suggest case studies for the VDAR analysis.

We conducted a case study with two epidemiological domain experts.
%
Both of them are practicing clinicians with focus on epidemiological research.
%
The first clinician is an specialist in internal medicine and also the leader of the SHIP.
%
One clinician is an radiologist and responsible for the image acquisition of the SHIP and has a deep understanding on the available image data.
%
\paragraph{Setup.} To bypass the large geographical distance between us and the domain experts and to take their busy schedule into account, the case study was done completely web based.
%
Since one major reason for the focus on web technologies, no software needed to be exchanged.
%
A website link was send to the users, which they could open with their browser of choice and start using the tool.
%
To make the user input comprehensible, we observed their monitor via screen sharing.
%
Communication was enabled via webcam supported voice over ip.
%
The total setup time including installing the screen sharing application was about five minutes.
%
Video-recordings of the sessions made it possible to evaluate the sessions later on.
\com{TODO: Name der Screensharing Software mit angeben (Teamviewer)?}
\paragraph{Procedure.}
At first, controlled mouse and keyboard of the clinicians pc and demonstrated the basic functionalities of the system.
%
This included the Heatmap, the correlation view, how to introduce parameter visualization, how the color coding of differences works and the pivot tables.
%
As they understood the concepts, we handed over the mouse and keyboard control and only observed from this point on.
%
The epidemiologists were given two tasks: one hypothesis-free analysis of the data and one starting with an assumption.

% \subsection{Exploratory Analysis of the Lumbar Spine Data Set}
\subsection{Hypothesis-free analysis}
Analyzing the data set without prior hypothesis requires a starting point giving an overview over the data first \cite{Shneiderman1996}.
%
With our tool, there are two ways to achieve this.
%
Performing a \emph{local investigation} by clustering all subjects or use the contingency matrix for a \emph{multivariate analysis}.
%
The latter was chosen by both domain experts.
%
They were excited by the possibility to look at all variables in the context of each other.
%
They stated that it "illumunates the black box of the data", making it possible to look at the data in an unbiased way.
%
Due to the dense mapping of information in the contingency matrix it was suggested to make this visualization full screen.
%
The radiologist was looking for correlations with shape related features in the data, finding that they correlate strong with each other as well as with leg pain, age, body height and hormone replacement therapy.
%

After this initial overview, both experts performed a \emph{multivariate analysis} by introducing features, such as age, waist circumference, weight or lumbar spine centerline curvature as bar charts views into the canvas area and selected subgroups to see how they are distributed and if they could observe unusual behavior in the mean-shapes.
%

% TODO: Contingency Matrix or Adjecency Matrix
\com{00.44 Clustering-Ergebnis Weiblich und Arbeitspensum Alter}

\subsection{Hypothesis-based analysis}
The radiologist performed a \emph{local investigation} by dragging the age variable into the canvas area, observing that age-bins 36-52 years and 52-68 years are nearly identical to the global mean, while 20-36 years and 68-84 years differ in size.
%
Performing a \emph{multivariate analysis} by selecting only female subjects via the additionally added gender feature it was observed, that distance was closest to subjects of 68-84 years old and largest to 20-36 years old.
%
The expert argued that this is most likely due to the shrinking of the spine with increasing age, making it more similar to the female average spine which is smaller than the global average \com{(00.29)}.
%

A second analysis was carried out on body weight.
%\com{Video 00.29. Analysis of Gender and Age - older subjects are closer to females, younger to males. This is due to shrinking of the spine in the aging process.}
\com{00.46 Second: Analysis of Female subjects via clustering, selction of cluster 2, analysis using Body weight and Age. Clustering Result if 1 gives large distance between other subjects. 00.49 How are can the differences be explained?}
\com{00.46 Feedback Body-Size/Body weight. Some features are sparse, suggestion was made to use quantile/quintile, resulting in equally occupied results. This also depends on what the user is intersted in - for clinical use the outliers may be interesting (57.30). (58.30) From perspective of a radiologists, quartile and quintiles make sense when establishing reference values for diseases; clinical relevant parameter, einflussfaktoren oder risikofaktoren zu identifizieren, je nach dem welche hypothese man verwendet.}
\com{Backpain three month > age. There are more subjects reporting backpain than those who dont. Since influences heavily the average shape - since most people have back pain, the average shape already is associated with back pain, which makes it difficult to extract parameters which reliably describe back pain. (1.05) The radiologist pointed out that for the mosaic view it would be beneficial to abstract the shape and curvature and categorize it into shapes s, i form. It was also pointed out that in the plot back pain>age, back pain can not be differenciated because the difference is mainly encoded via the age, more variables need to be added (gender, use of pain medications. this would be good for future work, providing a better way to subdivide the data set and construct groups. The presented method is very good to gain an overview).}
\subsection{Lessons Learned}
\com{ToDo: Noch einmal auf die VDAR Results eingehen}
%
The radiologist pointed out the unusual strong similarity of the L3 vertebrae throughout the population.
%
The medical explanation is that it represents a angular point of curvature of the lumbar spine.
%
A seconds explanation is the use of the L3 vertebra is initialization point of the lumbar spine model \com{(00.54)}.
%
The domain experts see another great potential in communicating insights efficiently using the presented visualizations \com{(1.08)}.
%
The radiologist pointed out that the image data representations would be better comparable if they would be abstracted into representative land marks, such as centers of the vertebrae and cardinal points of the lumbar spine canal.
%
Both experts emphasized the potential of the \emph{multivariate analysis} capabilities of the adjacency matrix for getting insight into a large amount of features simultaneously.
%
It is also useful to verify established but still controversial risk factors, such as the metabolic syndrome for coronary heart disease and whether the data set provides even more suitable risk factors.
%
Creating adjacency matrices for sub groups, such as different age bins can help to characterize the aging process by deriving age-specific risk factors.
%
%She also pointed out the benefits of an automatic categorization of the subjects into shape groups.%$$, which allows for even faster comparison.
\com{Employ a more abstract visualization of the spine to capture reliable information and enable better comparison between subjects}
%
%Since many large population studies incorporate medical image data this gains importance.
%\com{1.08 It was pointed out that this technique is also good for epidemiological publication because it is often more insigthfull to show these images instead of long tables. }
\com{ToDo Fazit über Web-Austausch, das Experten begeistert davon waren und dass es hervorragend funktioniert. Kleine Änderungen können sogar live programmiert werden ohne das etwas neu kompiliert werden muss (hier wäre ein Beispiel nett, z.B. Größe der Schrift irgendwo. Außerdem bringt der Browser viele Funktionen mit, wie z.B. die Suchfunktionen.)}

\com{Small User Interface Suggestions: Search-Bar for Subjects, making the elements of the contingency matrix selectable; See which Subject is currently selected}
\com{Enhancement contingency matrix: Display exact value of correlations, also information about distribution. More focus on this}
\com{Außerdem ein Anwendungsgebiet für die Matrix: Einschätzen metabolisches syndrom, eine Menge von Risikofaktoren die für Koronare Herzkrankheit Marker darstellen und die bereits länger in der Kritik sind. Hier kann getestet werden, ob es hierfür nicht bessere Indikatoren noch geben kann! (39.32)}
%\com{Also mosaic plots can be incorporated for several sub groups to see where the main differences are (for different age groups to characterize age)}
\com{1.20 More exact segmentations are useful since curvature changes are often the last change happening in a pathological process, after local deformations of the vertebrae}

% ---------------------------------------------------------------------------------------------------------------------------------------
% ---------------------------------------------------------------------------------------------------------------------------------------
% ---------------------------------------------------------------------------------------------------------------------------------------
\section{Old Application Section}
%
This section describes how our presented \emph{IVA} workflow is used for an epidemiological use case.
%
We applied the presented set of techniques to a data set compiled to analyze lower back pain. 
%
This is one of the most common reasons for an adult to see a physician in the Western civilization \cite{Backpain}.
%
Epidemiological analysis of lumbar back pain, such as the work of Harreby et al. \cite{Harreby1996}, is largely focused on non-image information.
%
In comparable studies, only a few shape-related features are included \cite{Lang2011}.
%
To our knowledge, this is the first approach, where shape-related information of the whole lumbar spine is analyzed along with other epidemiological features.
%
Determining risk factors in this area can lead to \cite{Fletcher2012}:
\begin{itemize}
	\item a better understanding of effects of preventive measures such as occupational health and safety regulations,
	\item prognostic features for diagnosis and treatment of lumbar back pain, and
	\item determination of particularly affected risk groups.
\end{itemize}
% Determining risk factors in this area can lead to \cite{Fletcher2012} (evtl. weglassen, Dopplung bei Epidemiological Background?):
% \begin{itemize}
% 	\item a better understanding of effects of preventive measures such as occupational health and safety regulations
% 	\item prognostic features for diagnosis and treatment of lumbar back pain
% 	\item determination of particularly effected risk groups
% \end{itemize}
%
Characterizing the healthy aging process of the spine is a long-term goal for determining age-normalized probabilities for spine-related diseases by incorporating individual risk factors.

@@

\section{Summary and Conclusion}
% TODO: Future work Bulletliste zusammenkürzen
In this paper, we presented an \emph{IVA} framework for the analysis of complex image-centric epidemiological data.
%
Hence, the framework allows for both, hypothesis-driven analysis and hypothesis generation.
%
The visualization of multivariate data using connected views and different views allows to get fast visual feedback about subject groups.
%
Brushing and linking makes the data tangible and adaptable to formulated hypotheses.
%
The use of pivot tables is familiar to epidemiologists while embracing the power of interactive adjustment of the shown features.
%
The automatic suggestion of correlations using contingency methods like \emph{Cram\'{e}r's V} triggers \emph{hypothesis generation} by highlighting features potentially overlooked by the experts.
%
Shape-based clustering assesses the variability of a anatomical structure in the context of non-image features such as disease indicators or lifestyle factors.
